---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load the downloaded data and convert to a data.table.  Set the **interval** column as a key.

```{r, echo=TRUE}
library(data.table)

activity1 <- data.table(read.csv('./data/activity.csv', header=TRUE))
setkey(activity1, interval)
numDays <- length(unique(activity1$date))
```

`r nrow(activity)` observations were found in the downloaded data.   
`r numDays` unique dates are present.

```{r}
summary(activity1)
```

***

## What is mean total number of steps taken per day?

Aggregate the activity data by date to sum the number of steps.

```{r, echo=TRUE}
stepsPerDay1 <- aggregate(. ~ date, data=activity1, FUN=sum)[,1:2]
hist(stepsPerDay1$steps, breaks=numDays, main="Steps Per Day", xlab="Steps", col="blue")

meanStepsPerDay1 <- mean(stepsPerDay1$steps)
medianStepsPerDay1 <- median(stepsPerDay1$steps)
```

#### Mean number of steps per day : `r format(meanStepsPerDay1, big.mark=',')`
#### Median number of steps per day : `r format(medianStepsPerDay1, big.mark=',')`

***

## What is the average daily activity pattern?

Aggregate the activity data by interval to find the mean value of the number of steps.

```{r, echo=TRUE}
avgStepsPerInterval <- data.table(aggregate(. ~ interval, data=activity1, FUN=mean)[,1:2])
setkey(avgStepsPerInterval, interval)

library(lattice)
xyplot(steps ~ interval, data=avgStepsPerInterval, type='l', xlab="Interval", ylab="Steps", main="Average Steps Per Interval")
```

***

## Inputing missing values

```{r, echo=TRUE}
numMissingSteps1 <- length(which(is.na(activity1$steps)))
```

`r numMissingSteps1` observation are missing values for number of steps. Create a duplicate dataset from the original data.  
Set missing step values equal to the mean interval value.

```{r, echo=TRUE ,warning=FALSE}
activity2 <- data.table(read.csv('../../data/activity.csv', header=TRUE))
setkey(activity2, interval)
activity2[avgStepsPerInterval, steps := ifelse(is.na(steps), i.steps, steps), nomatch=0]

numMissingSteps2 <- length(which(is.na(activity2$steps)))

stepsPerDay2 <- aggregate(. ~ date, data=activity2, FUN=sum)[,1:2]
hist(stepsPerDay2$steps, breaks=numDays, col="blue")

meanStepsPerDay2 <- mean(stepsPerDay2$steps)
medianStepsPerDay2 <- median(stepsPerDay2$steps)

pctDiffMeanStepsPerDay <- abs((meanStepsPerDay2 - meanStepsPerDay1) / meanStepsPerDay2) * 100
pctDiffMedStepsPerDay <- abs((medianStepsPerDay2 - medianStepsPerDay1) / medianStepsPerDay2) * 100

hist(stepsPerDay2$steps, breaks=numDays, main="Steps Per Day", xlab="Steps", col="blue")
```

#### Mean number of steps per day : `r format(meanStepsPerDay2, big.mark=',')` (`r format(pctDiffMeanStepsPerDay, digits=4)` % difference)  
#### Median number of steps per day : `r format(medianStepsPerDay2, big.mark=',')` (`r format(pctDiffMedStepsPerDay, digits=4)` % difference) 

***

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor (day.type) in activity2 signifying whether the date is a weekday or weekend.  
Calculated the mean number of steps per day.type and interval.

```{r, echo=TRUE}
activity2$date <- as.Date(activity2$date)
activity2[,day.type := ifelse(weekdays(date) %in% c("Saturday","Sunday"), "weekend","weekday")]

library(plyr)
avgStepsPerIntervalDayType <- ddply(activity2, c("day.type","interval"), avgSteps = mean(steps))

xyplot(avgSteps ~ interval | day.type, data=avgStepsPerIntervalDayType, layout=c(1,2), type='l', xlab="Interval", ylab="Steps", main="Average Steps Per Interval")
```





